---
title: "Introduction to causalgeex"
author: "Brian G. Barkley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to causalgeex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# `causalgeex` for estimating causal effects with empirical sandwich variance estimator

Currently implemented as of version 0.0.0.9000:
- Estimators of the average treament effect
    - IPTW
    - G-formula
    - Doubly Robust estimator
    
## Why use `causalgeex`? Easy implementation with valid inference!

- This package is inteded to be a "one-stop-shop" with emphasis on ease of use. 
- The estimators here are consistent and asymptotically normal following usual regularity assumptions.


### What's `geex`? 

[geex](https://cran.r-project.org/web/packages/geex/index.html) is a package designed for easy implementation of estimating equations. The `causalgeex` package is powered by `geex`

### What are estimating equations?

See [Stefanski and Boos (2002)](http://www.jstor.org/stable/3087324?seq=1#page_scan_tab_contents)

## How to install

The package can be installed from Github:

```{r, eval = FALSE}
devtools::install_github("BarkleyBG/causalgeex")
```

## Examples

Whenver possible, the causal effect estimate is listed last.

### Estimating the Average Treatment Effect

First, some data:

```{r}
library(causalgeex)

n <- 200
data <- data.frame(
  Covar1 = rnorm(n),
  Covar2 = rnorm(n)
)
trtprobs <- plogis(0.2  + 0.5*data$Covar1 + 0.1*data$Covar2 + 0.2*data$Covar1 * data$Covar2)
data$BinaryTrt <- rbinom(n, 1, trtprobs)

outprobs <- plogis(0.5  + 0.2*data$Covar1 - 0.2*data$Covar2 + 0.2*data$Covar1 * data$Covar2 -0.2*data$BinaryTrt)
data$BinaryOutcome <- rbinom(n, 1, outprobs)
```

#### Defensive Programming

Until rigorous tests and defensive programming are added, users are recommended to use the following coding schemes:

- Binary treatments should be indicator variables, with 1 for "treated" and 0 for "untreated". Use integers, not factors.
- Binary outcomes should be indicator variables, with for "outcome" and 0 for "no outcome". Use integers, not factors.


### Inverse Probability (of treament) Weighting

```{r}

```

### G-formula

### Doubly-Robust Estimation 

- Here we specify two formulas, one for each model. 
- For illustration we also pass in `deriv_control = geex::setup_deriv_control(method="simple")`. This results in quicker (but probably less accurate) derivatives in the variance computations. This is passed directly to the `geex` guts.

```{r}
outcome_regression_formula <- BinaryOutcome ~ BinaryTrt + Covar1*Covar2
treatment_model_formula <- BinaryTrt ~ Covar1 * Covar2

DRIPTW <- estimateDRIPTW(
  data = data,
  outcome_formula = outcome_regression_formula,
  treatment_formula = treatment_model_formula,
  outcome_model_method = "logistic",
  treatment_model_method = "logistic",
  deriv_control = geex::setup_deriv_control(method="simple")
)

(ests <- DRIPTW@estimates)
(vcov <- DRIPTW@vcov)
```

An estimate and 95\% confidence interval is then

```{r}
ests[10]
ests[10] + stats::qnorm(c(0.025,0.975))*vcov[10,10]
```




